% Acknowledgements ============================================================

\pdfbookmark[1]{Acknowledgments}{acknowledgments}
\chapter*{Acknowledgments}

As the saying goes, good premises do not entail good stories. Yet, this
dissertation would certainly not have come to its successful conclusion
without the help, support and trust of colleagues, friends and family.

First and foremost, I would like to sincerely thank my advisor Pierre Geurts
for his help, guidance and for the freedom I was granted throughout these
years.

I am grateful to all members of the jury for their interest in this work
and for taking the  time to evaluate this dissertation.

In alphabetical order, I would also like to thank my colleagues who all
contributed to create and maintain a pleasant and stimulating working
environment: Antonio, Arnaud, Benjamin, Damien, Fabien, Julien, Lo\"{i}c,
Louis, Marie, Olivier, Rapha\"{e}l, Van Anh, Vincent. Special thanks go to
Antonio, Arnaud and Vincent who accepted to proofread parts of this manuscript.

I want take this opportunity to thank the Scikit-Learn team and all its
contributors. This experience within the open source world really contributed
to shape my vision of science and software development towards a model
of rigor, pragmatism and openness. Thanks go to Ga\"{e}l, Olivier, Lars,
Mathieu, Andreas, Alexandre and Peter.

Special thanks go to the rowing team of the RCAE, for their friendship
and good mood in all circumstances. Guys, I thank you all.

Even if I never succeeded to fully explain my research topics, I would finally
like to warmly thank my dear friend J\'er\^{o}me and my family for their help
in moments of doubt.

Last but not least, Laura, I am forever grateful for your unconditional support
and love.
